The Unicode Cookbook \linebreak for Linguists

### 2018-06-18 book/chapters/implementation.tex
numerous edits

First, in Section \ref{installing-python-and-r} we explain how to install Python\footnote{\url{https://www.python.org/}} and R.\footnote{\url{https://www.r-project.org/}} Then in Sections \ref{python-implementations} and \ref{r-implementation}, we discuss our Python and R software packages, respectively. In addition to the material presented here to get users started, we maintain several case studies online that illustrate how to use orthography profiles in action. For convenience, we make these recipes available as Jupyter Notebooks\footnote{\url{http://jupyter.org/}} in our GitHub repository.\footnote{\url{https://github.com/unicode-cookbook/}} In the final section in this chapter, we also briefly describe a few recipes that we do not go into detail in this book.

@@

\noindent We set the default as the Unicode replacement character \uni{fffd}, \texttt{replacement character}.\footnote{\url{http://www.fileformat.info/info/unicode/char/fffd/index.htm}} But you can simply change this by specifying the replacement character when you load the orthography profile with the tokenizer.

@@

Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) \texttt{Shiny} webapps.

The bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-like machine. The exact location of these
executables is best found after installation of R the packages. The
location can be found by the following command in R.

@@

These different profiles can also be produced using the bash
executable \texttt{writeprofile} (see above for how to install the bash executable). 
This example is also included in the help file of the executable.

>>>>

First, in Section \ref{installing-python-and-r} we explain how to install Python\footnote{\url{https://www.python.org/}} and R.\footnote{\url{https://www.r-project.org/}} Then in Sections \ref{python-implementations} \& \ref{r-implementation}, we discuss our Python and R software packages, respectively. In addition to the material presented here to get users started, we maintain several case studies online that illustrate how to use orthography profiles in action. For convenience, we make these recipes available as Jupyter Notebooks\footnote{\url{http://jupyter.org/}} in our GitHub repository.\footnote{\url{https://github.com/unicode-cookbook/}} In the final section in this chapter, we also briefly describe a few recipes that we do not go into detail in this book.

@@

\noindent We set the default as the Unicode \texttt{replacement character} \uni{fffd}.\footnote{\url{http://www.fileformat.info/info/unicode/char/fffd/index.htm}} But you can simply change this by specifying the replacement character when you load the orthography profile with the tokenizer.

@@

Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) \texttt{Bash}
executables and (ii) \texttt{Shiny} webapps.

The Bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-like machine. The exact location of these
executables is best found after installation of R the packages. The
location can be found by the following command in R.

@@

These different profiles can also be produced using the bash
executable \texttt{writeprofile} (see above for how to install the Bash executable). 
This example is also included in the help file of the executable.

### 2018-06-18 book/chapters/ipa_meets_unicode.tex
numerous edits

There is one (minor) aspect of ordering for which the Unicode Standard does
present a canonical solution. Fortunately, this is uncontroversial from a
linguistic perspective. Diacritics in the Unicode Standard (i.e.~Combining
Diacritical Marks, see Section~\ref{pitfall-different-notions-of-diacritics})
are classified in so-called \textsc{canonical combining classes}. In practice,
the diacritics are distinguished by their position relative to the base
character.\footnote{See
\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for
a detailed description.} When applying a Unicode normalization (NFC or NFD, see
Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
positions are put in a specified order. This process therefore harmonizes the
difference between different encodings in some situations, for example in the
case of an extra-short creaky voice vowel <ḛ̆>. This grapheme cluster can be
encoded either as <e>+<\dia{0306}>+<\dia{0330}> or as
<e>+<\dia{0330}>+<\dia{0306}>. To prevent this twofold encoding, the Unicode
Standard specifies the second ordering as canonical (namely, diacritics
below are put before diacritics above).

@@

\begin{itemize}
	\item Symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
       because the latter were the symbols used by nearly all Khoisanists and
       Bantuists.
	\item The Americanist tradition of using using <\dia{030C}>, a
       \textsc{combining caron} at \uni{030C} for all postalveolar sounds, like
       in <š~ž~č~ǰ>, was not adopted because the Association members at the
       convention ``were not sufficiently impressed by arguments ... to the
       effect that these sounds formed a natural class, and thus is would be
       appropriate to recognize this by maintaining a common aspect to their
       symbolism'' \citep[62]{Ladefoged1990a}. 
\end{itemize}

@@

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode Standard, which is probably best considered an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive sucking sound similar
      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ejective or creaky voicing are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.\ the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

>>>>

There is one (minor) aspect of ordering for which the Unicode Standard does
present a canonical solution. Fortunately, this is uncontroversial from a
linguistic perspective. Diacritics in the Unicode Standard (i.e.~Combining
Diacritical Marks, see Section~\ref{pitfall-different-notions-of-diacritics})
are classified in so-called \textsc{canonical combining classes}. In practice,
the diacritics are distinguished by their position relative to the base
character.\footnote{For a detailed description, see: 
\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values}.} 
When applying a Unicode normalization (NFC or NFD, see
Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
positions are put in a specified order. This process therefore harmonizes the
difference between different encodings in some situations, for example in the
case of an extra-short creaky voice vowel <ḛ̆>. This grapheme cluster can be
encoded either as <e>+<\dia{0306}>+<\dia{0330}> or as
<e>+<\dia{0330}>+<\dia{0306}>. To prevent this twofold encoding, the Unicode
Standard specifies the second ordering as canonical (namely, diacritics
below are put before diacritics above).

@@

\begin{itemize}
	\item Symbols for clicks were changed from <ʇ~ʖ~ʗ> to <ǀ~ǁ~ǃ>
       because the latter were the symbols used by nearly all Khoisanists and
       Bantuists.
	\item The Americanist tradition of using using <\dia{030C}>, a
       \textsc{combining caron}\\ at \uni{030C} for all postalveolar sounds, like
       in <š~ž~č~ǰ>, was not adopted because the Association members at the
       convention ``were not sufficiently impressed by arguments ... to the
       effect that these sounds formed a natural class, and thus is would be
       appropriate to recognize this by maintaining a common aspect to their
       symbolism'' \citep[62]{Ladefoged1990a}. 
\end{itemize}

@@

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode Standard, which is probably best considered an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at \\
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive sucking sound similar
      to a fricative, for example <kǃ͓ʰ>; second, frictionalized vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant \\ phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ejective or creaky voicing are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.\ the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

### 2018-06-18 book/chapters/orthography_profiles.tex
numerous edits

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode normalization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} and \ref{pitfall-homoglyphs-in-IPA}). 
Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated  for search and comparison.

>>>>

In (1), the string <tsʰṍ̰shi> has been tokenized into ten Unicode code points
(using NFD normalization), delimited here by space. Unicode normalization is
required because sequences of code points can differ in their visual and logical
orders. For example, <õ̰> is ambiguous to whether it is the sequence of <o> +
<\dia{0303}> + <\dia{0330}> or <o> + <\dia{0330}> + <\dia{0303}>. Although these two variants are visually homoglyphs,
computationally they are different (see Sections \ref{pitfall-ipa-homoglyphs} \& \ref{pitfall-homoglyphs-in-IPA}). 
Unicode normalization should be applied to
this string to reorder the code points into a canonical order, allowing the data
to be treated  for search and comparison.

### 2018-06-18 book/chapters/pitfalls.tex
numerous edits

To understand why any problems arise it is important to realize that there are
basically three different approaches to font rendering. The most widespread is
Adobe's and Microsoft's \textsc{OpenType} system. This approach makes it
relatively easy for font developers, but the font itself does not include all
details about the precise placement of individual characters. For those details,
additional script descriptions are necessary. All such systems can lead to
unexpected behavior.\footnote{For more details about OpenType, see
\url{http://www.adobe.com/products/type/opentype.html} and
\url{http://www.microsoft.com/typography/otspec/}. Additional systems for
complex text layout are, among others, Microsoft's DirectWrite
\url{https://msdn.microsoft.com/library/dd368038.aspx} and the open-source
project HarfBuzz \url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}.}
Alternative systems are \textsc{Apple Advanced Typography} (AAT) and the
open-source \textsc{Graphite} system produced and maintained by the Non-Roman Script Initiative of SIL International 
(SIL).\footnote{More information about AAT can be found at
\url{https://developer.apple.com/fonts/}. \newline Graphite is described
in detail at:
\url{http://scripts.sil.org/default}.}
In these systems, a larger burden is placed on the description inside
the font.

@@

In general, finding a specific character in the Unicode Standard is often non-trivial. 
The names of the blocks can help, but they are not (and were never supposed
to be) a foolproof structure. It is neither the goal nor the aim of the Unicode
Consortium to provide a user interface to the Unicode Standard. If one often
encounters the problem of needing to find a suitable character, there are
various other useful services for end-users available.\footnote{The Unicode
website offers a basic interface to the code charts at
\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
particularly like PopChar from Ergonis Software, available for both Mac and
Windows. There are also various free websites that offer search interfaces
to the Unicode code tables, like \url{http://unicode-search.net} or
\url{http://unicode-search.net}. A further useful approach for searching characters
using shape matching is \url{http://shapecatcher.com}.}

@@

The actual intended meaning of a Unicode code point is a combination of the
name, the block and the character properties (see
Chapter~\ref{the-unicode-approach}). Further details about the underlying intentions 
with which a character should be used
are only accessible by perusing the actual decisions of the Unicode Consortium.
All proposals, discussions and decisions of the Unicode Consortium are publicly
available. Unfortunately there is not (yet) any way to easily find everything
that is ever proposed, discussed and decided in relation to a specific
code point of interest, so many of the details are often somewhat
hidden.\footnote{All proposals and other documents that are the basis of Unicode
decisions are available at \url{http://www.unicode.org/L2/all-docs.html}. The
actual decisions that make up the Unicode Standard are documented in the minutes
of the Unicode Technical Committee, available at
\url{http://www.unicode.org/consortium/utc-minutes.html}.}

>>>>

To understand why any problems arise it is important to realize that there are
basically three different approaches to font rendering. The most widespread is
Adobe's and Microsoft's \textsc{OpenType} system. This approach makes it
relatively easy for font developers, but the font itself does not include all
details about the precise placement of individual characters. For those details,
additional script descriptions are necessary. All such systems can lead to
unexpected behavior.\footnote{For more details about OpenType, see
\url{http://www.adobe.com/products/type/opentype.html} and
\url{http://www.microsoft.com/typography/otspec/}. Additional systems for
complex text layout are, among others, Microsoft's DirectWrite
(\url{https://msdn.microsoft.com/library/dd368038.aspx}) and the open-source
project HarfBuzz (\url{http://www.freedesktop.org/wiki/Software/HarfBuzz/}).}
Alternative systems are \textsc{Apple Advanced Typography} (AAT) and the
open-source \textsc{Graphite} system produced and maintained by the Non-Roman Script Initiative of SIL International 
(SIL).\footnote{More information about AAT can be found at:
\url{https://developer.apple.com/fonts/}. \newline Graphite is described
in detail at:
\url{http://scripts.sil.org/default}.}
In these systems, a larger burden is placed on the description inside
the font.

@@

In general, finding a specific character in the Unicode Standard is often non-trivial. 
The names of the blocks can help, but they are not (and were never supposed
to be) a foolproof structure. It is neither the goal nor the aim of the Unicode
Consortium to provide a user interface to the Unicode Standard. If one often
encounters the problem of needing to find a suitable character, there are
various other useful services for end-users available.\footnote{The Unicode
website offers a basic interface to the code charts at:
\url{http://www.unicode.org/charts/index.html}. As a more flexible interface, we
particularly like PopChar from Ergonis Software, available for both Mac and
Windows. There are also various free websites that offer search interfaces
to the Unicode code tables, like \url{http://unicode-search.net} or
\url{http://unicode-search.net}. Another useful approach for searching for characters
using shape matching \citep{Belongie2002} is: \url{http://shapecatcher.com}.}

@@

The actual intended meaning of a Unicode code point is a combination of the
name, the block and the character properties (see
Chapter~\ref{the-unicode-approach}). Further details about the underlying intentions 
with which a character should be used
are only accessible by perusing the actual decisions of the Unicode Consortium.
All proposals, discussions and decisions of the Unicode Consortium are publicly
available. Unfortunately there is not (yet) any way to easily find everything
that is ever proposed, discussed and decided in relation to a specific
code point of interest, so many of the details are often somewhat
hidden.\footnote{All proposals and other documents that are the basis of Unicode
decisions are available at: \url{http://www.unicode.org/L2/all-docs.html}. The
actual decisions that make up the Unicode Standard are documented in the minutes
of the Unicode Technical Committee, available at:
\url{http://www.unicode.org/consortium/utc-minutes.html}.}

### 2018-06-18 book/chapters/practical_recommendations.tex
remove extra curly
numerous edits

This chapter is meant to be a short guide for novice users who are not interested in the programmatic aspects presented in Chapters \ref{orthography-profiles} and \ref{implementation}. Instead, we provide links to quickly find general information about the Unicode Standard and the International Phonetic Alphabet (IPA). We target ordinary working linguists who want to know how to easily insert special characters into their digital documents and applications.

@@

\begin{itemize}
	\item \url{https://software.sil.org/doulos/}}
	\item \url{https://software.sil.org/charis/}
\end{itemize}

>>>>

This chapter is meant to be a short guide for novice users who are not interested in the programmatic aspects presented in Chapters \ref{orthography-profiles} \& \ref{implementation}. Instead, we provide links to quickly find general information about the Unicode Standard and the International Phonetic Alphabet (IPA). We target ordinary working linguists who want to know how to easily insert special characters into their digital documents and applications.

@@

\begin{itemize}
	\item \url{https://software.sil.org/doulos/}
	\item \url{https://software.sil.org/charis/}
\end{itemize}

### 2018-06-18 book/chapters/preface.tex
numerous edits

This text is meant as a practical guide for linguists, and programmers, who work with data in multilingual computational environments. We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.

The intersection of the Unicode Standard and the International Phonetic Alphabet is often not met without frustration by users. Nevertheless, the two standards have provided language researchers with a consistent computational architecture needed to process, publish and analyze data from many different languages. We bring to light common, but not always transparent, pitfalls that researchers face when working with Unicode and IPA.

Our research uses quantitative methods to compare languages and uncover and clarify their phylogenetic relations. However, the majority of lexical data available from the world's languages is in author- or document-specific orthographies. Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we created a suite of open-source Python and R tools to work with languages using profiles that adequately describe their orthographic conventions. Using orthography profiles and these tools allows users to segment text, analyze it, identify errors, and to transform it into different written forms.

We welcome comments and corrections of this text, our source code, and the use case studies that are supplement to this book. Please use the issue tracker or email us directly.

>>>>

This text is meant as a practical guide for linguists and programmers, who work with data in multilingual computational environments. We introduce the basic concepts needed to understand how writing systems and character encodings function, and how they work together.

The intersection of the Unicode Standard and the International Phonetic Alphabet is often met with frustration by users. Nevertheless, the two standards have provided language researchers with a consistent computational architecture needed to process, publish and analyze data from many different languages. We bring to light common, but not always transparent, pitfalls that researchers face when working with Unicode and IPA.

We use quantitative methods in our research to compare languages to uncover and clarify their phylogenetic relationships. However, the majority of lexical data available from the world's languages is in author- or document-specific orthographies. Having identified and overcome the pitfalls involved in making writing systems and character encodings syntactically and semantically interoperable (to the extent that they can be), we created a suite of open-source Python and R software packages to work with languages using profiles that adequately describe their orthographic conventions. Using orthography profiles and these tools allows users to tokenize and transliterate text from diverse sources, so that they can be meaningfully compared and analyzed.

We welcome comments and corrections regarding this book, our source code, and the supplemental case studies that we provide online.\footnote{\url{https://github.com/unicode-cookbook/}} Please use the issue tracker, email us directly, or make suggestions on PaperHive.\footnote{\url{https://paperhive.org/}}

### 2018-06-18 book/chapters/unicode_approach.tex
numerous edits

Each version of the Unicode Standard (\citealp{Unicode2018}, as of writing at
version 11) consists of a set of specifications and guidelines that include (i) a
core specification, (ii) code charts, (iii) standard annexes and (iv) a
character database.\footnote{All documents of the Unicode Standard are available
at \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
of terminology inside the Unicode Standard, their glossary is particularly
useful, available at \url{http://www.unicode.org/glossary/}. For a general
introduction to the principles of Unicode, Chapter 2 of the core specification,
called \textsc{general structure}, is particularly insightful. Unlike 
many other documents in the Unicode Standard, this general introduction is
relatively easy to read and illustrated with many interesting examples from
various orthographic traditions from all over the world.} The \textsc{core
specification} is a book aimed at human readers that describes the formal
standard for encoding multilingual text. The \textsc{code charts} provide a
human-readable online reference to the character contents of the Unicode
Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
are a set of technical standards that describe the implementation of the Unicode
Standard for software development, web standards, and programming languages.
The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
files that describe the character properties, including a set of rich character
and writing system semantics, for each character in the Unicode Standard. In
this section, we introduce the basic Unicode concepts, but we will leave out
many details. Please consult the above-mentioned full documentation for a more
detailed discussion. Further note that the Unicode Standard is exactly that,
namely a standard. It normatively describes notions and rules to be followed. In
the actual practice of applying this standard in a computational setting, a
specific implementation is necessary. The most widely used implementation of the
Unicode Standard is the \textsc{International Components for Unicode (ICU)},
which offers C/C++ and Java libraries implementing the Unicode
Standard.\footnote{More information about the ICU is available here:
\url{http://icu-project.org}.}

@@

Each code point in the Unicode Standard is associated with a set of
\textsc{character properties} as defined by the Unicode character property
model.\footnote{The character property model is described in
\url{http://www.unicode.org/reports/tr23/}, but the actual properties are
described in \url{http://www.unicode.org/reports/tr44/}. A simplified overview
of the properties is available at
\url{http://userguide.icu-project.org/strings/properties}. The actual code
tables listing all properties for all Unicode code points are available at
\url{http://www.unicode.org/Public/UCD/latest/ucd/}.} Basically, those
properties are just a long list of values for each character. For example, code
point \uni{0047} has the following properties (among many others): 
\begin{itemize}
	\item Name: LATIN CAPITAL LETTER G 
	\item Alphabetic: YES 
	\item Uppercase: YES 
	\item Script: LATIN 
	\item Extender: NO 
	\item Simple\_Lowercase\_Mapping: 0067 
\end{itemize}

>>>>

Each version of the Unicode Standard (\citealp{Unicode2018}, as of writing at
version 11) consists of a set of specifications and guidelines that include (i) a
core specification, (ii) code charts, (iii) standard annexes and (iv) a
character database.\footnote{All documents of the Unicode Standard are available
at: \url{http://www.unicode.org/versions/latest/}. For a quick survey of the use
of terminology inside the Unicode Standard, their glossary is particularly
useful, available at: \url{http://www.unicode.org/glossary/}. For a general
introduction to the principles of Unicode, Chapter 2 of the core specification,
called \textsc{general structure}, is particularly insightful. Unlike 
many other documents in the Unicode Standard, this general introduction is
relatively easy to read and illustrated with many interesting examples from
various orthographic traditions from all over the world.} The \textsc{core
specification} is a book aimed at human readers that describes the formal
standard for encoding multilingual text. The \textsc{code charts} provide a
human-readable online reference to the character contents of the Unicode
Standard in the form of PDF files. The \textsc{Unicode Standard Annexes (UAX)}
are a set of technical standards that describe the implementation of the Unicode
Standard for software development, web standards, and programming languages.
The \textsc{Unicode Character Database (UCD)} is a set of computer-readable text
files that describe the character properties, including a set of rich character
and writing system semantics, for each character in the Unicode Standard. In
this section, we introduce the basic Unicode concepts, but we will leave out
many details. Please consult the above-mentioned full documentation for a more
detailed discussion. Further note that the Unicode Standard is exactly that,
namely a standard. It normatively describes notions and rules to be followed. In
the actual practice of applying this standard in a computational setting, a
specific implementation is necessary. The most widely used implementation of the
Unicode Standard is the \textsc{International Components for Unicode (ICU)},
which offers C/C++ and Java libraries implementing the Unicode
Standard.\footnote{More information about the ICU is available here:
\url{http://icu-project.org}.}

@@

Each code point in the Unicode Standard is associated with a set of
\textsc{character properties} as defined by the Unicode character property
model.\footnote{The character property model is described in
\url{http://www.unicode.org/reports/tr23/}, but the actual properties are
described in \url{http://www.unicode.org/reports/tr44/}. A simplified overview
of the properties is available at: 
\url{http://userguide.icu-project.org/strings/properties}. The actual code
tables listing all properties for all Unicode code points are available at: 
\url{http://www.unicode.org/Public/UCD/latest/ucd/}.} Basically, those
properties are just a long list of values for each character. For example, code
point \uni{0047} has the following properties (among many others): 
\begin{itemize}
	\item Name: LATIN CAPITAL LETTER G 
	\item Alphabetic: YES 
	\item Uppercase: YES 
	\item Script: LATIN 
	\item Extender: NO 
	\item Simple\_Lowercase\_Mapping: 0067 
\end{itemize}
